Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)

### 2025-05-21 02-antecedents.qmd
update interaction effects
review - not rendered

Conceptually, *market justice* has been discussed in the literature as a normative principle that legitimates the distribution of economic rewards based on individual merit and effort. It is possible to trace the concept to the understanding of Lane [-@lane_market_1986], who defines market justice as a system of "earned deserts," whereby individuals are seen as deserving of a determined distributive outcome due to their effort and skills. In contrast, political justice,  emphasizes principles of equality and need, which are often represented by the welfare state action through social policies. An important remark is that the principles of market justice rely on the assumption that markets are neutral and self-regulating arenas, where individuals are treated fairly because they face the same formal rules of engagement [@lane_market_1986]. Consequently, the legitimacy of market justice stems from the assumption that inequalities are not only inevitable but fair---so long as the rules are transparent and opportunities are open. In this way, market justice provides a moral justification for inequality by framing it as a necessary outcome of individual responsibility.

Empirical studies have shown different strategies for the study of market justice preferences. A common approach in the literature is to gauge attitudes toward the legitimacy of inequality in specific domains, especially when linked to income differences. This can be traced to the seminal work of Kluegel and Smith [-@kluegel_beliefs_1981] who assessed the normative foundations that explain public support for economic inequality. Over time, this approach has been extended beyond income to include other market-mediated outcomes, such as education, healthcare, and/or pensions. For example, @vondemknesebeck_are_2016 and Immergut and Schneider [-@immergut_it_2020] examine whether citizens consider it fair that individuals with higher incomes can access better healthcare, while Lee and Stacey [-@lee_fairness_2023] apply a similar method in the context of education in Australia. These studies usually rely on a survey item asking respondents to evaluate the fairness of income-based access to welfare services, allowing for comparing justice perceptions across different contexts. Similarly, comparative studies have also relied on these indicators to study the support for market-based distribution in welfare systems [@svallfors_political_2007;@lindh_public_2015]. More recently, Castillo et al. [-@castillo_socialization_2024] introduced a single-item composite measure of market justice to assess student attitudes toward income-based access to education, healthcare, and pensions in Chile. These empirical strategies all aim to capture the extent to which individuals accept inequality when framed as a reflection of market outcomes.

The study of market justice preferences has increasingly focused on how they are shaped by individuals' socioeconomic position, normative beliefs, and the institutional context in which they are embedded. Across the literature, there is empirical evidence suggesting that individuals in more advantaged socioeconomic positions---those with higher occupational class, income, and education---are more likely to support market justice principles [@svallfors_political_2007; @koos_moral_2019]. This tendency reflects not only material self-interest but also a broader moral economy, in which winners of the market system internalize justifications for the status quo. At the same time, political ideology also plays a role ---such as economically conservative values ---  where more right-wing individuals show higher support for meritocracy and more skepticism towards redistribution. This is particularly salient in countries with more restricted public provision of social services. For example, Castillo et al. [-@castillo_socialization_2024] in Chile and, Lee and Stacey [-@lee_fairness_2023] in Australia show that those with right-leaning individuals are more supportive of market-based distribution of welfare. Beyond individual characteristics, country-level institutions also play a central role. In liberal welfare regimes like those of the United States or the United Kingdom, market justice preferences are more widespread, while in coordinated or social-democratic regimes---such as Sweden or Germany---citizens are generally more critical of market-based inequalities [@immergut_it_2020; @lindh_public_2015]. These findings suggest that market justice is also shaped contextually, which is represented by the development of welfare institutions.

@@

Additionally, this study examines whether these relationships vary over time in response to significant social and political events. In particular, we explore whether the political outburst of 2019 and the subsequent constitutional processes in Chile have altered how perceptions of inequality and meritocracy shape market justice attitudes. We expect that after these critical events, perceptions of inequality may have a stronger negative effect on market justice preferences, reflecting increased societal questioning of market-based allocation mechanisms.

@@

$H1$: Higher perceived economic inequality is associated with higher market justice preferences.

>>>>

Conceptually, *market justice* has been discussed in the literature as a normative principle that legitimates the distribution of economic rewards based on individual merit. It is possible to trace the concept to the understanding of Lane [-@lane_market_1986], who makes a contrast between market justice and political justice. The author defines market justice as a system of "earned deserts," whereby individuals are seen as deserving of a determined distributive outcome due to their effort and skills. In contrast, political justice emphasizes principles of equality and need, which are often represented by the welfare state action through social policies. An important remark is that the principles of market justice rely on the assumption that markets are neutral and self-regulating arenas, where individuals are treated fairly because they face the same formal rules of engagement [@lane_market_1986]. Consequently, the legitimacy of market justice stems from the assumption that inequalities are not only inevitable but fair---so long as the rules are transparent and opportunities are open. In this way, market justice provides a moral justification for inequality by framing it as a necessary outcome of individual responsibility.

Empirical studies have shown different strategies for the study of market justice preferences. A common approach in the literature is to gauge attitudes toward the legitimacy of inequality in specific domains, especially when linked to income differences. This can be traced to the seminal work of Kluegel and Smith [-@kluegel_beliefs_1981] who assessed the normative foundations that explain public support for economic inequality. Over time, this approach has been extended beyond income to include other market-mediated outcomes, such as education, healthcare, and/or pensions. For example, @vondemknesebeck_are_2016 and Immergut and Schneider [-@immergut_it_2020] examine whether citizens consider it fair that individuals with higher incomes can access better healthcare, while Lee and Stacey [-@lee_fairness_2023] apply a similar method in the context of education in Australia. These studies usually rely on a survey item asking respondents to evaluate the fairness of income-based access to welfare services, allowing for comparing justice perceptions across different contexts. Similarly, comparative studies have also considered these indicators to study the support for market-based distribution in welfare systems [@svallfors_political_2007;@lindh_public_2015]. More recently, Castillo et al. [-@castillo_socialization_2024] introduced a single-item composite measure of market justice to assess student attitudes toward income-based access to education, healthcare, and pensions in Chile. These empirical strategies all aim to capture the extent to which individuals accept inequality when framed as a reflection of market outcomes.

The study of market justice preferences has increasingly focused on how they are shaped by individuals' socioeconomic position, normative beliefs, and the institutional context in which they are embedded. Across the literature, there is empirical evidence suggesting that individuals in more advantaged socioeconomic positions---those with higher occupational class, income, and education---are more likely to support market justice principles [@svallfors_political_2007; @koos_moral_2019]. This tendency reflects not only material self-interest but also a broader moral economy, in which winners of the market system internalize justifications for the status quo. At the same time, political ideology also plays a role ---such as economically conservative values ---  where more right-wing individuals show higher support for meritocracy and more skepticism towards redistribution. This is particularly salient in countries with more restricted public provision of social services. For example, Castillo et al. [-@castillo_socialization_2024] in Chile and, Lee and Stacey [-@lee_fairness_2023] in Australia show that those with right-leaning individuals are more supportive of market-based distribution of welfare. Beyond individual characteristics, country-level institutions also play a central role. In liberal welfare regimes like those of the United States or the United Kingdom, market justice preferences are more widespread, while in coordinated or social-democratic regimes---such as Sweden or Germany---citizens are generally more critical of market-based inequalities [@immergut_it_2020; @lindh_public_2015]. These findings suggest that market justice is also shaped contextually, which is represented by the development of welfare institutions, in line with the policy-feedbacks literature [campbell_policy_2012; pierson_when_1993].

@@

Additionally, this study examines whether these relationships vary over time in response to significant social and political events. In particular, we explore whether the political outburst of 2019 and the subsequent constitutional processes in Chile --- where socio-economic inequality was widely challenged --- have altered how perceptions of inequality and meritocracy shape market justice attitudes. We expect that after these critical events, perceptions of inequality may have a stronger negative effect on market justice preferences, reflecting increased societal questioning of market-based allocation mechanisms.

@@

$H1$: Higher perceived economic inequality is associated with less market justice preferences.

### 2025-05-21 03-methods.qmd
review - not rendered

**Market justice preferences**: The dependent variable in this study is preferences for market justice. This construct is operationalized through three items that capture how strongly individuals justify conditioning access to social services---healthcare, pensions, and education---on income. Specifically, the justification of inequality in healthcare is assessed by the question: "Is it fair in Chile that people with higher incomes can access better healthcare than people with lower incomes?" The same question is posed for pensions and education. In all cases, respondents indicate their level of agreement on a five-point Likert scale ranging from 1 ("strongly disagree") to 5 ("strongly agree"). Additionally, we include a composite measure of "market justice preferences", calculated as the average of these three items ($\alpha$ = 0.84). This index ranges from 1 to 5, with higher values indicating stronger preferences for market justice (see @tbl-summary1).

@@

**Perception of economic inequality**: The main independent variable refers to the perception of economic inequality, measured through the perceived wage gap. [@castillo_cual_2009; @hadler_why_2005; @gijsberts_thelegitimation_1999]. This measure is derived from the salary gap between the perceived salaries of jobs at opposite ends of the occupational hierarchy. Specifically, it relies on the division between the perceived salary of a large-company president and that of an unskilled worker [@Castillo2011]. Higher values of this term indicate a greater perception of economic inequality between occupations located at the extremes of the status continuum. This measure includes a logarithmic term in order to adjust income magnitudes (usually fewer cases with high income):

@@

```{r}
#| label: tbl-summary2
#| tbl-cap: "Independent variables for the first wave (2016)"
#| tbl-cap-location: top
#| results: asis
#| echo: false

@@

Sociodemographic and attitudinal variables are included to control for potential composition effects in the population. In terms of sociodemographic characteristics, we incorporate per capita household income quintile, educational level (1=Less than Universitary, 2=Universitary), age (in years), and sex (1=Male, 2=Female), which have been previously shown to influence market justice preferences significantly [@lindh_public_2015; @castillo_socialization_2024]. Regarding attitudinal variables, we include political identification (1=Left, 2=Center, 3=Right, 4=No identification) and subjective social status (ranging from 1 to 10) as they may confound the relationship between market justice preferences and perceptions of inequality and meritocracy [@schneider_poverty_2015].

>>>>

**Market justice preferences**: The dependent variable in this study is preferences for market justice. This construct is operationalized through three items that capture how strongly individuals justify conditioning access to social services---healthcare, pensions, and education--- basen on individual income. Specifically, the justification of inequality in healthcare is assessed by the question: "Is it fair in Chile that people with higher incomes can access better healthcare than people with lower incomes?" The same question is posed for pensions and education. In all cases, respondents indicate their level of agreement on a five-point Likert scale ranging from 1 ("strongly disagree") to 5 ("strongly agree"). Additionally, we include a composite measure of "market justice preferences", calculated as the average of these three items ($\alpha$ = 0.84). This index ranges from 1 to 5, with higher values indicating stronger preferences for market justice (see @tbl-summary1).

@@

**Perception of economic inequality**: The main independent variable refers to the perception of economic inequality, measured through the perceived wage gap [@castillo_cual_2009; @hadler_why_2005; @gijsberts_thelegitimation_1999]. This measure is derived from the salary gap between the perceived salaries of jobs at opposite ends of the occupational hierarchy. Specifically, it relies on the division between the perceived salary of a large-company president and that of an unskilled worker [@Castillo2011]. Higher values of this term indicate a greater perception of economic inequality between occupations located at the extremes of the status continuum. This measure includes a logarithmic term in order to adjust income magnitudes (usually fewer cases with high income):

@@

```{r}
#| label: tbl-summary2
#| tbl-cap: "Independent variables ELSOC survey (descriptives for first wave 2016)"
#| tbl-cap-location: top
#| results: asis
#| echo: false

@@

Sociodemographic and attitudinal variables are included to control for potential composition effects in the population. In terms of sociodemographic characteristics, we incorporate per capita household income quintile, educational level (1=Less than Universitary, 2=Universitary), age (in years), and sex (1=Male, 2=Female), which have been previously shown to influence market justice preferences significantly [@lindh_public_2015; @castillo_socialization_2024]. Regarding attitudinal variables, we include political identification (1=Left, 2=Center, 3=Right, 4=No identification) and subjective social status (ranging from 1 to 10) as they may affect the relationship between market justice preferences, perceptions of inequality, and meritocracy [@schneider_poverty_2015].

### 2025-05-21 04-analisis.qmd
review - not rendered

@fig-alluvial shows the annual frequencies of market justice preferences for healthcare, pensions, and education from 2016 to 2023. Each year presents stacked percentage frequencies, and the flows between them reflect opinion changes among the same individuals from one year to the next, given that we are using panel data. For instance, of the 40.8% who strongly disagreed with justifying inequality in healthcare in 2019, around 24.3% maintained that position in 2022, while the remaining 16.5% shifted toward other response categories—primarily moving into disagreement rather than strong disagreement. Overall, more than half of the respondents exhibit a high level of disagreement (disagree + strongly disagree) with inequality in these three social service areas over time. Despite this general pattern, recent waves show a slight decrease in disagreement and a corresponding rise in support for inequality. Specifically, in healthcare and education, although disagreement remains substantial, agreement (agree + strongly agree) increased from 7.4% and 7.2% in 2019 to 13.1% and 14.2% in 2023, respectively. This shift is most evident in pensions, where the combined agree/strongly agree category grew by about 10 percentage points, from 16.9% in 2016 to 28% in 2023.

@@

datos.health <- df_study1 %>% 
  mutate(just_health = factor(just_health, 
                              levels = c("Strongly agree",
                                         "Agree",
                                         "Neither agree nor desagree",
                                         "Desagree",
                                         "Strongly desagree"))) %>% 
  group_by(idencuesta, ola) %>% 
  count(just_health) %>% 
  group_by(ola) %>% 
  mutate(porcentaje=n/sum(n)) %>% 
  ungroup() %>% 
  na.omit() %>% 
  mutate(wave = case_when(ola == 1 ~ "2016",
                          ola == 2 ~ "2017",
                          ola == 3 ~ "2018",
                          ola == 4 ~ "2019",
                          ola == 5 ~ "2022",
                          ola == 6 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                      "2017",
                                      "2018",
                                      "2019",
                                      "2022",
                                      "2023")))

etiquetas.health <- df_study1 %>%
  mutate(just_health = factor(just_health, 
                              levels = c("Strongly agree",
                                         "Agree",
                                         "Neither agree nor desagree",
                                         "Desagree",
                                         "Strongly desagree"))) %>% 
  group_by(ola, just_health) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(ola) %>%
  mutate(porcentaje = count / sum(count)) %>% 
  na.omit() %>% 
  mutate(idencuesta = 1,
         wave = case_when(ola == 1 ~ "2016",
                          ola == 2 ~ "2017",
                          ola == 3 ~ "2018",
                          ola == 4 ~ "2019",
                          ola == 5 ~ "2022",
                          ola == 6 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                        "2017",
                                        "2018",
                                        "2019",
                                        "2022",
                                        "2023")))

@@

datos.pension <- df_study1 %>% 
   mutate(just_pension = factor(just_pension, 
                              levels = c("Strongly agree",
                                         "Agree",
                                         "Neither agree nor desagree",
                                         "Desagree",
                                         "Strongly desagree"))) %>% 
  group_by(idencuesta, ola) %>% 
  count(just_pension) %>% 
  group_by(ola) %>% 
  mutate(porcentaje=n/sum(n)) %>% 
  ungroup() %>% 
  na.omit() %>% 
  mutate(wave = case_when(ola == 1 ~ "2016",
                          ola == 2 ~ "2017",
                          ola == 3 ~ "2018",
                          ola == 4 ~ "2019",
                          ola == 5 ~ "2022",
                          ola == 6 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                        "2017",
                                        "2018",
                                        "2019",
                                        "2022",
                                        "2023")))

etiquetas.pension <- df_study1 %>%
  mutate(just_pension = factor(just_pension, 
                              levels = c("Strongly agree",
                                         "Agree",
                                         "Neither agree nor desagree",
                                         "Desagree",
                                         "Strongly desagree"))) %>% 
  group_by(ola, just_pension) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(ola) %>%
  mutate(porcentaje = count / sum(count)) %>% 
  na.omit() %>% 
  mutate(idencuesta = 1,
         wave = case_when(ola == 1 ~ "2016",
                          ola == 2 ~ "2017",
                          ola == 3 ~ "2018",
                          ola == 4 ~ "2019",
                          ola == 5 ~ "2022",
                          ola == 6 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                        "2017",
                                        "2018",
                                        "2019",
                                        "2022",
                                        "2023")))

@@

datos.educ <- df_study1 %>% 
  mutate(just_educ = factor(just_educ, 
                              levels = c("Strongly agree",
                                         "Agree",
                                         "Neither agree nor desagree",
                                         "Desagree",
                                         "Strongly desagree"))) %>%
  group_by(idencuesta, ola) %>% 
  count(just_educ) %>% 
  group_by(ola) %>% 
  mutate(porcentaje=n/sum(n)) %>% 
  ungroup() %>% 
  na.omit() %>% 
  mutate(wave = case_when(ola == 1 ~ "2016",
                          ola == 2 ~ "2017",
                          ola == 3 ~ "2018",
                          ola == 4 ~ "2019",
                          ola == 5 ~ "2022",
                          ola == 6 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                        "2017",
                                        "2018",
                                        "2019",
                                        "2022",
                                        "2023")))

etiquetas.educ <- df_study1 %>%
    mutate(just_educ = factor(just_educ, 
                              levels = c("Strongly agree",
                                         "Agree",
                                         "Neither agree nor desagree",
                                         "Desagree",
                                         "Strongly desagree"))) %>%
  group_by(ola, just_educ) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(ola) %>%
  mutate(porcentaje = count / sum(count)) %>% 
  na.omit() %>% 
  mutate(idencuesta = 1,
         wave = case_when(ola == 1 ~ "2016",
                          ola == 2 ~ "2017",
                          ola == 3 ~ "2018",
                          ola == 4 ~ "2019",
                          ola == 5 ~ "2022",
                          ola == 6 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                        "2017",
                                        "2018",
                                        "2019",
                                        "2022",
                                        "2023")))

@@

Regarding the main variables of this study, @fig-meanchange depicts their average changes over the years. We observe an increase in the average level of market justice preferences in the most recent waves, which begins in 2019. The highest average consistently appears for perceived economic inequality, although this variable shows a downward trend of roughly one point over time. Interestingly, while perceptions of economic inequality declined in the latest measurements (2022–2023), market justice preferences increased. The meritocracy measures remain stable, though the perception that individuals are rewarded for intelligence is slightly higher than the perception that they are rewarded for their effort.

@@

@tbl-interactions1 reports whether perceived economic inequality moderates the impact of meritocratic perceptions on market-justice preferences. The interaction terms in the within-person specification (Model 8 and Model 9) do not reach statistical significance. However, in the between-person specification (Model 10 and Model 11), interaction terms reach statistical significance, indicating that perceptions of economic inequality do meaningfully alter the effect of meritocracy on support for market allocation of social services. These effects are positive and mean that when the perception of inequality is greater, the positive effect of the meritocratic variables on market justice preferences is greater. These interaction effects are visualized in @fig-interact

>>>>

@fig-alluvial shows the annual frequencies of market justice preferences for healthcare, pensions, and education from 2016 to 2023. Each year presents stacked percentage frequencies, and the flows between them reflect opinion changes among the same individuals from one year to the next, given that we are using panel data. For instance, of the 40.8% who strongly disagreed with justifying inequality in healthcare in 2019, around 24.3% maintained that position in 2022, while the remaining 16.5% shifted toward other response categories—primarily moving into disagreement rather than strong disagreement. Overall, more than half of the respondents exhibit a high level of disagreement (disagree + strongly disagree) with inequality in these three social service areas over time. Despite this general pattern, recent waves show a slight decrease in disagreement and a corresponding rise in support for market-justice inequality. Specifically, in healthcare and education, although disagreement remains substantial, agreement (agree + strongly agree) increased from 7.4% and 7.2% in 2019 to 13.1% and 14.2% in 2023, respectively. This shift is most evident in pensions, where the combined agree/strongly agree category grew by about 10 percentage points, from 16.9% in 2016 to 28% in 2023.

@@

datos.health <- df_study1 %>% 
  mutate(just_health = factor(just_health, 
                              levels = c("Strongly agree",
                                         "Agree",
                                         "Neither agree nor disagree",
                                         "Disagree",
                                         "Strongly disagree"))) %>% 
  group_by(idencuesta, ola) %>% 
  count(just_health) %>% 
  group_by(ola) %>% 
  mutate(porcentaje=n/sum(n)) %>% 
  ungroup() %>% 
  na.omit() %>% 
  mutate(wave = case_when(ola == 1 ~ "2016",
                          ola == 2 ~ "2017",
                          ola == 3 ~ "2018",
                          ola == 4 ~ "2019",
                          ola == 5 ~ "2022",
                          ola == 6 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                      "2017",
                                      "2018",
                                      "2019",
                                      "2022",
                                      "2023")))

etiquetas.health <- df_study1 %>%
  mutate(just_health = factor(just_health, 
                              levels = c("Strongly agree",
                                         "Agree",
                                         "Neither agree nor disagree",
                                         "Disagree",
                                         "Strongly disagree"))) %>% 
  group_by(ola, just_health) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(ola) %>%
  mutate(porcentaje = count / sum(count)) %>% 
  na.omit() %>% 
  mutate(idencuesta = 1,
         wave = case_when(ola == 1 ~ "2016",
                          ola == 2 ~ "2017",
                          ola == 3 ~ "2018",
                          ola == 4 ~ "2019",
                          ola == 5 ~ "2022",
                          ola == 6 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                        "2017",
                                        "2018",
                                        "2019",
                                        "2022",
                                        "2023")))

@@

datos.pension <- df_study1 %>% 
   mutate(just_pension = factor(just_pension, 
                              levels = c("Strongly agree",
                                         "Agree",
                                         "Neither agree nor disagree",
                                         "Disagree",
                                         "Strongly disagree"))) %>% 
  group_by(idencuesta, ola) %>% 
  count(just_pension) %>% 
  group_by(ola) %>% 
  mutate(porcentaje=n/sum(n)) %>% 
  ungroup() %>% 
  na.omit() %>% 
  mutate(wave = case_when(ola == 1 ~ "2016",
                          ola == 2 ~ "2017",
                          ola == 3 ~ "2018",
                          ola == 4 ~ "2019",
                          ola == 5 ~ "2022",
                          ola == 6 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                        "2017",
                                        "2018",
                                        "2019",
                                        "2022",
                                        "2023")))

etiquetas.pension <- df_study1 %>%
  mutate(just_pension = factor(just_pension, 
                              levels = c("Strongly agree",
                                         "Agree",
                                         "Neither agree nor disagree",
                                         "Disagree",
                                         "Strongly disagree"))) %>% 
  group_by(ola, just_pension) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(ola) %>%
  mutate(porcentaje = count / sum(count)) %>% 
  na.omit() %>% 
  mutate(idencuesta = 1,
         wave = case_when(ola == 1 ~ "2016",
                          ola == 2 ~ "2017",
                          ola == 3 ~ "2018",
                          ola == 4 ~ "2019",
                          ola == 5 ~ "2022",
                          ola == 6 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                        "2017",
                                        "2018",
                                        "2019",
                                        "2022",
                                        "2023")))

@@

datos.educ <- df_study1 %>% 
  mutate(just_educ = factor(just_educ, 
                              levels = c("Strongly agree",
                                         "Agree",
                                         "Neither agree nor disagree",
                                         "Disagree",
                                         "Strongly disagree"))) %>%
  group_by(idencuesta, ola) %>% 
  count(just_educ) %>% 
  group_by(ola) %>% 
  mutate(porcentaje=n/sum(n)) %>% 
  ungroup() %>% 
  na.omit() %>% 
  mutate(wave = case_when(ola == 1 ~ "2016",
                          ola == 2 ~ "2017",
                          ola == 3 ~ "2018",
                          ola == 4 ~ "2019",
                          ola == 5 ~ "2022",
                          ola == 6 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                        "2017",
                                        "2018",
                                        "2019",
                                        "2022",
                                        "2023")))

etiquetas.educ <- df_study1 %>%
    mutate(just_educ = factor(just_educ, 
                              levels = c("Strongly agree",
                                         "Agree",
                                         "Neither agree nor disagree",
                                         "Disagree",
                                         "Strongly disagree"))) %>%
  group_by(ola, just_educ) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(ola) %>%
  mutate(porcentaje = count / sum(count)) %>% 
  na.omit() %>% 
  mutate(idencuesta = 1,
         wave = case_when(ola == 1 ~ "2016",
                          ola == 2 ~ "2017",
                          ola == 3 ~ "2018",
                          ola == 4 ~ "2019",
                          ola == 5 ~ "2022",
                          ola == 6 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                        "2017",
                                        "2018",
                                        "2019",
                                        "2022",
                                        "2023")))

@@

Regarding the main dependent and independent variables of this study, @fig-meanchange depicts their average changes over the years. We observe an increase in the average level of market justice preferences in the most recent waves, which begins in 2019. The highest average consistently appears for perceived economic inequality, although this variable shows a downward trend of roughly one point over time. Interestingly, while perceptions of economic inequality declined in the latest measurements (2022–2023), market justice preferences increased. The meritocracy measures remain stable, though the perception that individuals are rewarded for intelligence is slightly higher than the perception that they are rewarded for their effort.

@@

@tbl-interactions1 reports whether perceived economic inequality moderates the impact of meritocratic perceptions on market-justice preferences. Contrary to our hypothesis, the interaction terms in the within-person specification (Model 8 and Model 9) do not reach statistical significance. However, in the between-person specification (Model 10 and Model 11), interaction terms reach statistical significance, indicating that perceptions of economic inequality do meaningfully alter the effect of meritocracy on support for market allocation of social services. These effects are positive and mean that when the perception of inequality is greater, the positive effect of the meritocratic variables on market justice preferences is greater. These interaction effects are visualized in @fig-interact

### 2025-05-21 paper.qmd
review - not rendered

---
title: "Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)"
format:
  docx:
    number-sections: true
  html: 
    code-fold: true
    html-math-method: katex
    number-sections: true
  pdf: 
    geometry: margin=2cm
    template-partials: 
      - title.tex
    keep-tex: true
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    number-sections: true
editor: source
bibliography: "input/bib/merit-edjust.bib"
csl: "input/bib/apa6.csl"
author:
  - name: Juan Carlos Castillo
    corresponding: true
    orcid: 0000-0003-1265-7854
    email: juancastillov@uchile.cl
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 3
  - name: Andreas Laffert
    orcid: 0000-0002-9008-2454
    email: alaffertt@estudiante.uc.cl 
    affiliations:
      - ref: 4
  - name: Kevin Carrasco
    orcid: 0000-0002-9348-8361
    email: kevin.carrasco@ug.uchile.cl
    affiliations:
      - ref: 2
  - name: Julio Iturra
    orcid: 0000-0003-1993-3479
    email: julioiturrasanhueza@gmail.com 
    affiliations:
      - ref: 5     
affiliations: 
  - id: 1
    name: Departamento de Sociología, Universidad de Chile
  - id: 2
    name: Centro de estudios del conflicto y cohesión social (COES)
  - id: 3
    name: Núcleo milenio de desigualdades y oportunidades digitales (NUDOS)
  - id: 4
    name: Instituto de Sociología, Pontificia Universidad Católica de Chile
  - id: 5
    name: International Graduate School of Social Sciencies (BIGSSS), University of Bremen, Germany
citeproc: true
abstract: | 
  My abstract. \newline
  **Keywords**: meritocracy, social inequality, inequality justification, COVID-19
link-citations: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
colorlinks: true
fig-height: 4
fig-width: 7.5
header-includes:
  - \usepackage{threeparttable}
---

>>>>

---
title: "Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)"
format:
  html: 
    code-fold: true
    html-math-method: katex
    number-sections: true
  docx:
    number-sections: true
  pdf: 
    geometry: margin=2cm
    template-partials: 
      - title.tex
    keep-tex: true
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    number-sections: true
editor: source
bibliography: "input/bib/merit-edjust.bib"
csl: "input/bib/apa6.csl"
author:
  - name: Juan Carlos Castillo
    corresponding: true
    orcid: 0000-0003-1265-7854
    email: juancastillov@uchile.cl
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 3
  - name: Andreas Laffert
    orcid: 0000-0002-9008-2454
    email: alaffertt@estudiante.uc.cl 
    affiliations:
      - ref: 4
  - name: Kevin Carrasco
    orcid: 0000-0002-9348-8361
    email: kevin.carrasco@ug.uchile.cl
    affiliations:
      - ref: 2
  - name: Julio Iturra
    orcid: 0000-0003-1993-3479
    email: julioiturrasanhueza@gmail.com 
    affiliations:
      - ref: 5     
affiliations: 
  - id: 1
    name: Departamento de Sociología, Universidad de Chile
  - id: 2
    name: Centro de estudios del conflicto y cohesión social (COES)
  - id: 3
    name: Núcleo milenio de desigualdades y oportunidades digitales (NUDOS)
  - id: 4
    name: Instituto de Sociología, Pontificia Universidad Católica de Chile
  - id: 5
    name: International Graduate School of Social Sciencies (BIGSSS), University of Bremen, Germany
citeproc: true
abstract: | 
  This study investigates the relationship between perceptions of economic inequality, meritocratic beliefs, and preferences for market justice in Chile between 2016 and 2023. Using six waves of panel data from the Chilean Longitudinal Social Survey (ELSOC; N = 1,519), the analysis examines how subjective assessments of inequality shape attitudes toward the role of merit in access to key social services such as healthcare, education, and pensions. Results indicate that rising perceptions of inequality are associated with greater support for redistributive policies; however, individuals with strong meritocratic convictions are more likely to legitimize existing disparities. The study also considers the influence of major social movements during this period, which appear to have reshaped public discourse and perceptions of fairness. These findings contribute to a deeper understanding of how beliefs about justice and equity evolve in contexts marked by persistent inequality and entrenched market-oriented frameworks.
  \newline
  **Keywords**: Economic inequality, meritocracy, market justice, Chile, public preferences, inequality perception
link-citations: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
colorlinks: true
fig-height: 4
fig-width: 7.5
---
